Supplementary Materials from: \textit{Innovation and elaboration on the avian tree of life}; Science Advances, in review

### 2023-06-07 Manuscript/main.tex
Update from revisions

By studying the phylogenetic line of least resistance and related properties of the evolutionary rate matrix \cite{caetano2017ratematrix, caetano2018estimating, slater2019hierarchy} at different scales we aim to gain new insights into the macro- and mega-evolutionary trajectories that lead to the diversity of beak shapes among extant bird species and clades.
While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades.
The diversity of bird beaks has been well-studied at broad scales, particularly in relation to dietary ecology \cite{olsen2017feeding, felice2019, bright2016shapes, navalon2019, pigot2020macroevolutionary}.
Previous studies have assessed how variation in rates of evolution of beak shape has generated this diversity \cite{cooney2017mega,chira2018correlates}, revealing a general pattern of phylogenetically clustered evolutionary rates interspersed with major shifts deep in avian evolutionary history and in isolated lineages.
Major shifts indicate lineages that occupy distinct areas of beak shape space relative to their close relatives.
For example, a major shift in beak shape arises in the early divergence of Strisores separating hummingbirds (Trochilidae) from swifts (Apodidae), perhaps marking changes in ontogenetic trajectories \cite{navalon2021} as hummingbirds diverged from broad billed ancestors (e.g. the Eocene stem Trochilidae fossils \textit{Parargornis messelensis} and \textit{Eocypselus rowei} \cite{mayr2003new, ksepka2013fossil}). %@L_strisores_example
By assessing the multivariate directions of the evolution of beak shapes of species and clades we aim to deepen our understanding of how diversity in morphological form arises at macro- and mega-evolutionary scales in birds.
To do this we use data on bird beaks derived from 3D scans of museum specimens for $>$8700 species ($\sim$85\% of all extant bird species) and generate evolutionary rate matrices describing the directions of multivariate beak shape evolution.
The phylogenetic breadth of the data allows us to compare directions of beak evolution of subclades (e.g. orders) within the class Aves (the megaevolutionary scale) and of species within subclades (the macroevolutionary scale).%@L_macromega_definition2

@@

\subsection{Modeling nested trait covariance}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of beak shape variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves) (Fig.
 \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
To estimate the major axis of beak shape variation we estimate the evolutionary rate matrix \cite{Houle2017,Machado2020} by fitting Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods). %@L_ei_meth_precision.
To assess the roles of elaboration and innovation in beak shape evolution across scales we fit our pGLMMs as nested models in which we define evolutionary rate matrices, and therefore major axes of beak shape variation, for
i) the entire phylogeny of all species included in the data (hereafter the class-wide phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level), and
iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation across scales.

@@

Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses} - Fig. \ref{Fig:orthogonality}).
The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig. \ref{Fig:orthogonality}).
Comparisons of orientation am%@L_ei_through_time_plotsong subclades (i.e. orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig. \ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another.
This differs from previous inference from a subset of our data analysed  without incorporating phylogeny, that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
This suggests that there is no megaevolutionary analogue of the genetic line of least resistance and instead that megaevolutionary shifts more likely reflect changes in the adaptive landscape.
Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}. % @L_occasionaly

@@

\subsection{Species elaborate \textit{and} innovate, clades innovate} %TG: @L_TLDR_nuance
The disconnect between observations of reorientation of trait space at the macroevolutionary scale, and an apparent predominance of $elaboration_{species}$ could be viewed as a paradox.
We formalize this paradox by testing whether elaboration is more common at higher taxonomic levels (megaevolutionary scale) than at the species level (macroevolutionary scale).
We compared the area under the curve of the scaled density of $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
We then measured the difference between both areas and found that there was a significant difference in $innovation_{species}$ and $innovation_{clade}$ but no clear difference in $elaboration_{species}$ and $elaboration_{clade}$ (Fig. \ref{Fig:relative_EI}).
This difference confirms our observation that innovation is indeed common and dominant at the megaevolutionary scale, however, the contributions of elaboration and innovation are shown to be indistinguishable at the macroevolutionary scale.
So, although species often evolve preferentially along a shared phylogenetic major axes of beak variation within clades, there is frequent deviation from these major axes, and the phylogenetic major axes of beak variation change among clades.

@@

\textbf{Phylogenetic generalised least squares analysis.} We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
Beak shape divergence was calculated as the Euclidean distance of each species beak shape from the centroid of beak shape of all birds.
We then fitted a series of nested models with distance as the response and $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$, and $innovation_{clade}$ as predictors.
The most complex model included all two-way interaction terms:

>>>>

By studying the phylogenetic line of least resistance and related properties of the evolutionary rate matrix \cite{caetano2017ratematrix, caetano2018estimating, slater2019hierarchy} at different scales we aim to gain new insights into the macro- and mega-evolutionary trajectories that lead to the diversity of beak shapes among extant bird species and clades.
While there is no strict definition that separates macro- and mega-evolution, we define the macroevolutionary scale as species divergence within clades and the megaevolutionary scale as divergence of clades within larger clades. %@L_mega_macro_def
The diversity of bird beaks has been well-studied at broad scales, particularly in relation to dietary ecology \cite{olsen2017feeding, felice2019, bright2016shapes, navalon2019, pigot2020macroevolutionary}.
Previous studies have assessed how variation in rates of evolution of beak shape has generated this diversity \cite{cooney2017mega,chira2018correlates}, revealing a general pattern of phylogenetically clustered evolutionary rates interspersed with major shifts deep in avian evolutionary history and in isolated lineages.
Major shifts indicate lineages that occupy distinct areas of beak shape space relative to their close relatives.
For example, a major shift in beak shape arises in the early divergence of Strisores separating hummingbirds (Trochilidae) from swifts (Apodidae), perhaps marking changes in ontogenetic trajectories \cite{navalon2021} as hummingbirds diverged from broad billed ancestors (e.g. the Eocene stem Trochilidae fossils \textit{Parargornis messelensis} and \textit{Eocypselus rowei} \cite{mayr2003new, ksepka2013fossil}). %@L_strisores_example
By assessing the multivariate directions of the evolution of beak shapes of species and clades we aim to deepen our understanding of how diversity in morphological form arises at macro- and mega-evolutionary scales in birds.
To do this we use data on bird beaks derived from 3D scans of museum specimens for $>$8700 species ($\sim$85\% of all extant bird species) and generate evolutionary rate matrices describing the directions of multivariate beak shape evolution.
The phylogenetic breadth of the data allows us to compare directions of beak evolution of subclades (e.g. orders) within the class Aves (the megaevolutionary scale) and of species within subclades (the macroevolutionary scale).%@L_macromega_definition2

@@

\subsection{Modeling nested trait covariance}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires: 1) large, multivariate datasets to allow exploration of trait covariances at different scales, 2) reliable and efficient computational methods to estimate the major axes of beak shape variation, and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the macroevolutionary scale (species within orders) and megaevolutionary scale (orders within the class Aves) (Fig.
 \ref{Fig:cheatsheet} and \ref{Fig:mcmcmcglmm}).
To estimate the major axis of beak shape variation we estimate the evolutionary rate matrix \cite{Houle2017,Machado2020} by fitting Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable \cite{MCMCglmm}.
The major axis can be identified from the leading eigenvector of the posterior distribution of evolutionary rate matrices.
We regard the major axis of the evolutionary rate matrix as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{steppan2002comparative, robinson2013quantifying} and posit that it represents the among species line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
This major axis of beak variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
We can then estimate elaboration and innovation based on that major axis: elaboration is defined as the position on the phylogenetic major axis of beak variation and innovation is defined as the distance from the phylogenetic major axis of beak variation (see Methods). %@L_ei_meth_precision.
To assess the roles of elaboration and innovation in beak shape evolution across scales we fit our pGLMMs as nested models in which we define evolutionary rate matrices, and therefore major axes of beak shape variation, for
i) the entire phylogeny of all species included in the data (hereafter the class-wide phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level), and
iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation across scales.

@@

Our observations of heterogeneity in the orientation of the phylogenetic major axis of beak variation among clades is further supported by consistent evidence for high amounts of orthogonality of clades relative to the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses} - Fig. \ref{Fig:orthogonality}; note that these comparisons are based on the evolutionary rate matrix). %@L_orientation_phylo
The median angle of the phylogenetic major axis of beak variation for subclades approaches orthogonality, differing from the class-wide phylogenetic major axis of beak variation by $68.14^\circ$ ($95$\% CI: $22.83^\circ$-$89.09^\circ$ - Fig. \ref{Fig:orthogonality}).
Comparisons of orientation among subclades (i.e. orders within super-orders) show similar differences (median = $47.14^\circ$; $95$\% CI: $13.64^\circ$-$87.62^\circ$ - Fig. \ref{Fig:orthogonality}), suggesting that reorientations in trait space are largely unconstrained at the megaevolutionary scale and are no more likely to occur along any one axis than another. 
This differs from previous inference from a subset of our data analysed  without incorporating phylogeny, that implied generally consistent and low dimensionality within clades \cite{cooney2017mega}.
This suggests that there is no megaevolutionary analogue of the genetic line of least resistance and instead that megaevolutionary shifts more likely reflect changes in the adaptive landscape.
Across scales there is remarkable flexibility in the routes to innovation, consistent with the idea that morphological divergence may be less constrained in deep-time than is sometimes assumed \cite{venditti2011multiple}. % @L_occasionaly

@@

\subsection{Species elaborate \textit{and} innovate, clades innovate} %TG: @L_TLDR_nuance
The disconnect between observations of reorientation of trait space at the macroevolutionary scale, and an apparent predominance of $elaboration_{species}$ could be viewed as a paradox.
We formalize this paradox by testing whether elaboration is more common at higher taxonomic levels (megaevolutionary scale) than at the species level (macroevolutionary scale).
We compared the area under the curve of the scaled density of $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ and $innovation_{clade}$.
We then measured the difference between both areas and found that there was a significant difference in $innovation_{species}$ and $innovation_{clade}$ but no clear difference in $elaboration_{species}$ and $elaboration_{clade}$ (Fig. \ref{Fig:relative_EI}).
This difference confirms our observation that innovation is indeed common and dominant at the megaevolutionary scale, however, the contributions of elaboration and innovation are shown to be indistinguishable at the macroevolutionary scale.
So, although species often evolve preferentially along a shared phylogenetic major axes of beak variation within clades, there is frequent deviation from these major axes, and the phylogenetic major axes of beak variation change among clades.
This could be explained in part because the evolutionary rate matrix (\textbf{R}) is more influenced by shifts in the adaptive landscape through time and thus reflecting natural selection in deeper nodes more than in younger nodes (Fig. \ref{Fig:figure_ei_clade_through_time} and \ref{Fig:figure_ei_species_through_time}). %@L_R_selection

@@

\textbf{Phylogenetic generalised least squares analysis.} %@L_pgls_methods
We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
Beak shape divergence was calculated as the Euclidean distance of each species beak shape from the centroid of beak shape of all birds.
We then fitted a series of nested models with distance as the response and $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$, and $innovation_{clade}$ as predictors.
The most complex model included all two-way interaction terms:
